A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-28 tsibble.Rmd
section 3

The choice of tidy representation of temporal data arises from a data- and model-oriented perspective, which can accommodate all of the operations that are to be performed on the data in time-based contexts. Figure \@ref(fig:tsibble-fit) marks where this new abstraction is placed in the "tidy" model, which is referred to as a "tsibble". Adhering to the "tidy data" principles, in tsibble:

@@

To materialize the abstraction of the tsibble, a subset of tuberculosis cases estimated by @tb-data presented in Table \@ref(tab:tb-sub) is used as an example. It contains 12 observations and 5 variables landing in a tidy data form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. From tidy data to tsibble data, "index" and "key" should be declared: column `year` as the "index" variable, and column `country` together with `gender` as the "key" variable forming the observational units. Column `count` is the only measured variable in this data, but the data structure is sufficiently flexible to hold more measurements; for example, slotting the corresponding population size (if known) into the data column for normalizing the count later. Note this data further satisfies the must-have for the distinct rows determined by index and key, and regularly spans over one-year interval.

@@

Time provides the contextual basis for temporal data. Time can be seen in numerous representations, from sequential numerics to the most commonly accepted date-times. Regardless of this diversity, time should be inherently ordered from past to present, so should the index variable to a tsibble.

Index is an explicit data variable rather than a masked attribute (such as in the `ts` and `zoo` classes), exposing a need for more accessible and transparent time operations. It is often to visualize and model seasonal effects of measurements of interest, meaning that time components such as time of day and day of week from the index should be easily extracted upon request. But when the index is available only as meta information, an obstacle is inevitably created for analysts to write such simple queries in an ad hoc and programmatic manner. From an analytical point of view it should be discouraged.

@@

What subjects/entities to be observed over time leads to the second component of a tsibble--key. Key can consist of empty, single, or multiple variables identifying units measured along the way. When only a single observational unit is present in the table, key can be left with no variables to be specified. However, when multiple units exist in the data, key should be supplied by identifying variables to sufficiently define the units. In longitudinal data, key can be thought of as "panel" (such as in the Stata) but constrained to a single variable in existing data structures. In tsibble, key allows for multiple variables of nesting, crossing, or union relations [@wilkinson2006grammar], that can be useful for forecasting reconciliation [@fpp; @R-hts] and richer visualization. For example, Table \@ref(tab:tb-sub) describes the number of tuberculosis cases for each gender across the countries every year. This suggests that the key comprises at least columns `gender` and `country`. Since `country` is nested within `continent`, it is a free variable to be included in the key specification.

@@

Inspired by "primary key" [@codd_relational_1970], a unique identifier for each observation in a relational database, tsibble key also uniquely identifies each observational unit over time. When constructing a tsibble, any duplicates of key-index (i.e. key-value) pairs will fail, because duplicates signal a data quality issue, which would likely affect subsequent analyses and hence decision making. For example, either `gender` or `country` alone is not enough to be the key for the tuberculosis data. Analysts are encouraged to better understand the data, or reason about the process of data cleaning when handling duplicates. Figure \@ref(fig:tsibble-tidy) peels the "tidy" module with clear routes required for a tsibble. The rigidity of tsibble, as the fundamental data infrastructure, warrants the validity of temporal data analysis for the later stage.

@@

The principal divide of temporal data is regularly versus irregularly spaced data. Event data typically involves irregular time intervals, such as flight schedules or customer transactions. This type of data can flow into event-based data modeling, but would need to be processed or regularized to fit models that expect data with a fixed-time interval. When data indexed in regular time space, tsibble computes the greatest common divisor as the time interval from the index, or an unknown interval if one observation is available for each unit.

By extending one of the "tidy data" rules, "Each type of observational units should form a table", each observational unit shares a common interval in a tsibble. Although tsibble reports one single interval despite data of fixed or mixed intervals, data of mixed intervals should be organized in separate tsibbles instead for well-tailored analysis and models.

This tiny piece of information will be carried over for tsibble-centric functions, for example to make implicit missing time handling handy and be in harmony with statistics and models that use seasonal periods.

>>>>

The choice of tidy representation of temporal data arises from a data- and model-oriented perspective, which can accommodate all of the operations that are to be performed on the data in time-based contexts. Figure \@ref(fig:tsibble-fit) marks where this new abstraction is placed in the "tidy" model, which is referred to as a "tsibble". Adapting the "tidy data" principles, in tsibble:

@@

To materialize the abstraction of the tsibble, a subset of tuberculosis cases [@tb-data], as presented in Table \@ref(tab:tb-sub), is used as an example. It contains 12 observations and 5 variables landing in a tidy data form. Each observation comprises the number of people who are diagnosed with tuberculosis for each gender at three selected countries in the years of 2011 and 2012. From tidy data to tsibble data, "index" and "key" should be declared: column `year` as the "index" variable, and column `country` together with `gender` as the "key" variable forming the observational units. Column `count` is the only measured variable in this data, but the data structure is sufficiently flexible to hold more measurements; for example, slotting the corresponding population size (if known) into the data column for normalizing the count later. Note, this data further satisfies the must-have for the distinct rows determined by index and key, and regularly spans over one-year intervals.

@@

Time provides the contextual basis for temporal data. Time can be seen in numerous representations, from sequential numerics to the most commonly accepted date-times. Regardless of this diversity, time should be inherently ordered from past to present, so should be the index variable to a tsibble.

Index is an explicit data variable rather than a masked attribute (such as in the `ts` and `zoo` classes), exposing a need for more accessible and transparent time operations. It is often necessary to visualize and model seasonal effects of measurements of interest, meaning that time components, such as time of day and day of week, should be easily extracted from the index. When the index is available only as meta information, it provides an obstacle for analysts by complicating the writing of even simple queries, often requiring special purpose programming. From an analytical point of view this should be discouraged.

@@

What subjects/entities are to be observed over time, leads to the second component of a tsibble--key. The key can consist of empty, single, or multiple variables identifying units measured along the way. When only a single observational unit is present in the table, no key needs to be specified. However, when multiple units exist in the data, the key should be supplied by identifying variables to sufficiently define the units. In longitudinal data, the key can be thought of as "panel" (such as in the Stata) but constrained to a single variable in existing data structures. In tsibble, the key allows for multiple variables of nesting, crossing, or union relations [@wilkinson2006grammar], that can be useful for forecasting reconciliation [@fpp; @R-hts] and richer visualization. For example, Table \@ref(tab:tb-sub) describes the number of tuberculosis cases for each gender across the countries every year. This suggests that the key comprises at least columns `gender` and `country`. Since `country` is nested within `continent`, continent can be included in the key specification, but is not needed.

@@

Inspired by a "primary key" [@codd_relational_1970], a unique identifier for each observation in a relational database, the tsibble key also uniquely identifies each observational unit over time. When constructing a tsibble, any duplicates of key-index (i.e. key-value) pairs will fail, because duplicates signal a data quality issue, which would likely affect subsequent analyses and hence decision making. For example, either `gender` or `country` alone is not enough to be the key for the tuberculosis data. Analysts are encouraged to better understand the data, or reason about the process of data cleaning when handling duplicates. Figure \@ref(fig:tsibble-tidy) peels the "tidy" module with clear routes required for a tsibble. The rigidity of tsibble, as the fundamental data infrastructure, warrants the validity of temporal data analysis for the later stage.

@@

The principal divide of temporal data is regularly versus irregularly spaced data. Event data typically involves irregular time intervals, such as flight schedules or customer transactions. This type of data can flow into event-based data modeling, but would need to be processed, or regularized, to fit models that expect data with a fixed-time interval.

There are three possible interval types: a number, unknown, and irregular. To determine the interval for regularly spaced data, tsibble computes the greatest common divisor, and is a number. If only one observation is available for each unit, which may occur after aggregating data, the interval is reported as unknown. When the data arrives with irregular time, like event data, the interval would be specified as irregular, to prevent the tsibble creator attempting to guess an interval.

To abide by the "tidy data" rules -- "Each type of observational units should form a table" -- in a tisbble each observational unit shares a common interval. This means that a tsibble will report one single interval, whether the data has a fixed or mixed set of intervals. To handle mixed interval data, it should be organized into separate tsibbles for a well-tailored analysis.

This tiny piece of information, the interval, is carried over for tsibble-centric operations. For example, this makes implicit missing time handling convenient, and harmoniously operates with statistical calculations, and models, on seasonal periods.
